---
title: "Activity- Scatterplot Matrix and Heat Maps"
author: "Gordon Anderson"
output: html_document
---

### Scatterplot Matrix

A quick way to view the relationships between several columns in a data set is by generating a scatterplot matrix. This is easily produced with a call to plot in basic R.

The data are results from a Java programming course survey including the final exam scores.
CLASS_LVL: 1st-year, 2nd-year, 3rd-year, 4th-year
GENDER: female, male
PREV: previous programming experience- none, some, java
MAJOR: ChemBiochem, Biology, OtherLifeSci, GeoPhysicalSci, EnvironmentalSci, Engineering, CompSci, Math, Humanities, SocialSci, Business, Undeclared
LECATTEND: How often did you attend lecture: Rarely, Sometimes, Mostly
DISCATTEND: How often did you attend discussion section: Rarely, Sometimes, Mostly
FINAL: Final exam score [0,100]

First, read in the data.
```{r}
cd.df<- read.csv("courseData.csv")
summary(cd.df)
```

Clean the data by omitting the NAs and set the factor levels for the survey columns.

```{r}
cd.df<-na.omit(cd.df)

cd.df$CLASS_LVL<-factor(cd.df$CLASS_LVL)	
levels(cd.df$CLASS_LVL)<- c("1yr","2yr","3yr","4yr")

cd.df$GENDER<-factor(cd.df$GENDER)
levels(cd.df$GENDER)<- c("female","male")

cd.df$PREV<-factor(cd.df$PREV)
levels(cd.df$PREV)<- c("none","some","java")

cd.df$MAJOR<-factor(cd.df$MAJOR)
levels(cd.df$MAJOR)<- c("ChemBio","Bio","OthLifeSci","Geo","Eng","CmpSci","Math","Hum","SocSci","Bus","Und")

cd.df$LECATTEND<-factor(cd.df$LECATTEND)	
levels(cd.df$LECATTEND)<- c("Rarely","Sometimes","Mostly")

cd.df$DISCATTEND<-factor(cd.df$DISCATTEND)	
levels(cd.df$DISCATTEND)<- c("Rarely","Sometimes","Mostly")

summary(cd.df)
```

Generate a scatterplot matric in basic R with plot. This can take a minute or two.

```{r}
plot(cd.df)
```

May not be very readable. You can try this with subsets of the data to get a better picture, but let's try a library that will produce a more decipherable result.
Scatterplot meatrices are a quick way to visualize relationships between variables in the data set.

```{r}
library(GGally)

ggpairs(cd.df, columns=c("GENDER", "CLASS_LVL", "PREV", "FINAL"),
    diag=list(continuous="density",   discrete="bar"), axisLabels="show")

ggpairs(cd.df, columns=c("GENDER", "CLASS_LVL", "FINAL"),
    diag=list(continuous="density",   discrete="bar"), axisLabels="show")

ggpairs(cd.df, columns=c("LECATTEND", "DISCATTEND", "FINAL"),
    diag=list(continuous="density", discrete="bar"), axisLabels="show")

ggpairs(cd.df, columns=c("MAJOR", "FINAL"),
    diag=list(continuous="density", discrete="bar"), axisLabels="show")
```

### Heat Map

A heatmap is a way to visualize a table of numerical data. Instead of numbers, colored cells are displayed to represent discrete levels of values. This is a quick way to view relationships between rows and columns in the data set.

The data set consists of crime statistics for each state in the USA.

```{r}
cr.df<- read.csv("crimeRatesByState.csv")
summary(cr.df)
```

Now a few steps to get the data ready for the heatmap plot.

```{r}
# Create row names- use the first column.
row.names(cr.df) <- cr.df$state

#Next, remove the first column as it is not needed.
cr.df <- cr.df[,-1]

#Transform the data frame to a matrix for the heatmap plot.
cr.matrix <- data.matrix(cr.df)
```

Now call the heatmap function. The "col" parameter cntrols the color palette.

```{r}
cr.heatmap <- heatmap(cr.matrix, Rowv=NA, Colv=NA, col = heat.colors(256), scale="column", margins=c(5,10))
```

This library allows for more color options.

```{r}
library(RColorBrewer)

cr.heatmap <- heatmap(cr.matrix, Rowv=NA, Colv=NA, col = brewer.pal(9, "Blues"), scale="column", margins=c(5,10))
```

You can sort the data on a column to see if it makes any patterns of interest more apparent.

```{r}
cr.df <- cr.df[order(cr.df$murder),]

cr.matrix <- data.matrix(cr.df)

cr.heatmap <- heatmap(cr.matrix, Rowv=NA, Colv=NA, col = brewer.pal(9, "Blues"), scale="column", margins=c(5,10))
```
